---
title: "itsar-vignette"
author: "William M. Old, Kristofor Webb"
date: "6/11/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{itsar-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  fig.width=7, fig.height=7,
  fig.align = "center"
  )
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

## Using the itsar package to identify protein targets of staurosporine from a protein lysate experiment  

This vignette demonstrates how to use the **itsar** package to analyze thermal profiling proteomic data collected using 
the isothermal shift assay (iTSA) method as published in Ball, KA et al. 2020. This method 
provides improved experimental design and statistical power for identifying the binding targets of small
molecules in complex proteomes, in either lysates or cells.


You can see more documentation of the itsar package once installed by typing at the command prompt `?itsar`.
Once there, click on the "Index" hyperlink at the bottom of page to see a list of functions, 
each with documentation.

iTSA Paper:
Ball KA, Webb KJ, Coleman SJ, Cozzolino KA, Jacobsen J, Jones KR, Stowell MHB, Old WM (2020).
An isothermal shift assay for proteome scale drug-target identification. Communications Biology 3, 1–10.
DOI: 10.1038/s42003-020-0795-6

## Load required packages for this vignette
Load R packages, including the itsar package (make sure it is installed from CU Boulder Old lab GitHub)
```{r loading-libs, message=FALSE}
library(itsar)
library(tidyverse)
require(magrittr)
require(dplyr)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.11")
# requireNamespace("limma", quietly=T)
# requireNamespace("sva", quietly=T)
BiocManager::install(c("limma", "sva"))


options(stringsAsFactors = FALSE)

```

Ensure that sample names and data column names are correct by reading in the header for the proteinGroups.txt file.
In this example, we are analyzing data from the iTSA paper that performed an iTSA experiment at 52C using 
the broad spectrum kinase inhibitor, staurosporine. MaxQuant was used to quantify the Drug vs. Control TMT10 experiment
in which drug and vehicle conditions alternated channels across the 10-plex (see paper for more details). 
The MaxQuant proteinGroups.txt file contains the protein-level TMT channel intensities that have been corrected for
channel-specific isotope impurities, with protein annotation information as well. We only focus on the 52C experiment here.

```{r}
#Read proteinGroups file
proteinGroupFile <- system.file("extdata",
                                       "proteinGroups_itsa_staurosporine_3temp_20190402KB.txt.gz",
                                       package = "itsar")
# read the first line of file to retrieve the column names, for properly setting up the sample names
# gzfile() is used because we are reading a gzipped tab delimited file
# 
pgfilecolnames <- as.character(
   read.table(
      file = gzfile(proteinGroupFile),
      header = F,
      nrows = 1,
      sep = "\t",
      na.strings = c("NA","NaN","Infinite"),
      fill = T,
      stringsAsFactors = F,
      quote = "\"")[1,])
 
### Filter data columns & define TMT data columns
# identifies column numbers that contain Label corrected intensity data (TMT lot-specific correction)
tmtdata_colnames <- grep("Reporter intensity corrected \\d+ 52C", pgfilecolnames, value = T)

```

Use the `itsar::read_maxquant_proteingroupfile()` function to read in the TMT protein data into a data frame
and convert to tibble (see dplyr package for more information on tibbles and tidyverse. This function returns 
a list containing a data frame of the protein TMT data and a character vector of column names.

```{r}

plist <- itsar::read_maxquant_proteingroupfile(
  proteinGroups.filename = proteinGroupFile,
  datacolnames = tmtdata_colnames,
  uniprot.id.column = "Majority protein IDs",
  uniprot.acc.column = "Fasta headers",
  null.zeros = T,
  log2.transform = F,
  is.gzip.file = T)

prot.table <- plist$prot.table
tmtdata_colnames <- plist$prot.colnms

tmtdata_colnames
prot.table <- as_tibble(prot.table)

```

Remove poorly quantified proteins with less than 2 unique peptides, using the dplyr function `dplyr::filter`

```{r}
# Remove poorly quantified proteins with less than 2 unique peptides
prot.table <- dplyr::filter(prot.table, Reverse != "+", 
                            Potential.contaminant != "+", 
                            Only.identified.by.site != "+",
                             Reporter.intensity.count.3.52C > 2,
                            Unique.peptides > 1)



```


Check for duplicate gene symbols (there will be) and make unique names using the annotation in the 
"Gene.names" column as primary names and the annotation in "Protein.IDs" as name for those that do not have an gene name. 
Then check uniqueness

```{r}

# Are there any duplicated gene names?

prot.table$Gene.names %>% duplicated() %>% any()

prot.table %>% dplyr::group_by(Gene.names) %>% 
  dplyr::summarize(frequency = n()) %>% 
  dplyr::arrange(desc(frequency)) %>% 
  dplyr::filter(frequency > 1)

#prot.table <- DEP::make_unique(prot.table, "Gene.names", "Protein.IDs", delim = ";")

# prot.table %>% dplyr::group_by(name) %>% 
#   dplyr::summarize(frequency = n()) %>% 
#   dplyr::arrange(desc(frequency)) %>% 
#   dplyr::filter(frequency > 1)

```

Check the densities of TMT data, will be lognormally distributed if columns haven't been log2 transformed
```{r}
limma::plotDensities(prot.table[,tmtdata_colnames], legend = "topright")

```

Log2 transform and quantile normalize 
```{r}
prot.table[,tmtdata_colnames] <- log2(prot.table[,tmtdata_colnames] )

limma::plotDensities(prot.table[,tmtdata_colnames], legend = "topright")

prot.table[,tmtdata_colnames] <- limma::normalizeBetweenArrays(as.matrix(prot.table[,tmtdata_colnames]),
                                                        method="quantile" )


```

Set up the targets matrix, which describes the experimental design. We are going to compare two models, 
one in which we correct for unexplained batch effects, and one in which we don't. 
First set up two different model matrices

```{r}


targets <- data.frame(samples = tmtdata_colnames,
                      treatment = factor(rep(c("Vehicle","Drug"), 5), levels = c("Vehicle","Drug")))
targets

origdesign <- model.matrix(~0+treatment, data = targets)
colnames(origdesign) <- c("Vehicle","Drug")  #V=Vehicle; D=Drug

origdesign

design <- model.matrix(~treatment, data = targets)
colnames(design) <- c("intercept","Drug") 


```
## Statistical testing

Run model without the batch effect correction (empirical Bayes linear modeling) 

```{r}
### Use limma for empirical bayes


fit <- limma::lmFit(prot.table[,tmtdata_colnames], origdesign)

fit <- limma::eBayes(fit) ##Apply empirical Bayes smoothing to the standard errors.

contrast <- limma::makeContrasts("Drug-Vehicle", levels=origdesign)

fit2 <- limma::contrasts.fit(fit, contrast)
fit2$Amean <- log2(prot.table$Intensity.52C)

fit2 <- limma::eBayes(fit2, trend = TRUE)
```

Use the get_toptable() function `itsar::get_toptable()`  (from itsar package)

```{r}
pstats <- itsar::get_toptable(fit2,
                              coef = 1, 
                              glist = prot.table[,c("id", "Gene.names", "uniprot.id")], 
                              suffix = ".uncor" )

pstats <- pstats %>% arrange(P.Value.uncor)

sum(pstats$P.Value.uncor <= 0.05)
sum(pstats$P.Value.uncor <= 0.01)
sum(pstats$P.Value.uncor<= 0.001)

pstats

```

Run model that incorporates latent variables that correct batch effects using
the R package sva. sva creates surrogate variables that model batch effects in 
the data, which are then added to the specified variables used in the 
empirical Bayes linear modeling with limma. 

```{r}

mod0 <- model.matrix(~1, data = targets)
colnames(mod0) <- "intercept"
dmat <- as.matrix(prot.table[,tmtdata_colnames])

numsv <- sva::num.sv(dmat, design, method = "be", B = 100)


svobj <- sva::sva(dmat,design,mod0,n.sv = 3) 

modSv = cbind(design,svobj$sv)
mod0Sv = cbind(mod0,svobj$sv)
pValuesSv = sva::f.pvalue(dmat,modSv,mod0Sv)
qValuesSv = p.adjust(pValuesSv,method="BH")
sum(qValuesSv < 0.1, na.rm = T)


svafit <- lmFit(dmat , modSv)
contrast.matrix <- cbind("intercept" = c(1,0, rep(0,svobj$n.sv) ), "Drug"= c(0,1,rep(0,svobj$n.sv)))


fitContrasts = limma::contrasts.fit(svafit,contrast.matrix)

svafit$Amean <- log2(prot.table$Intensity.52C)

fit2_sva <- eBayes(fitContrasts, trend = TRUE)
```

Use the itsar function get_toptable() to retrieve all tested proteins, and sort by p-value.
The p-values are corrected using the BH method.
```{r}
pstats_sva <- itsar::get_toptable(fit2_sva, coef = "Drug", 
                                  glist = prot.table[,c("id", "Gene.names", "uniprot.id")], 
                                  suffix = ".sva" )

pstats_sva <- pstats_sva %>% arrange(P.Value.sva)

sum(pstats_sva$adj.P.Val.sva <= 0.05)
sum(pstats_sva$adj.P.Val.sva <= 0.01)
sum(pstats_sva$adj.P.Val.sva<= 0.001)

pstats_sva

```

## Plotting

Make volcano plots to show the relationship between the effect size for drug-induced 
thermal-stabilization (log2(Staurosporine/DMSO)) and -log10(p-value) for differential expression 
Uses the itsar function `itsar::ggvolcano()`

```{r}

pstats$logp <- -log10(pstats$P.Value.uncor)
pstats_sva$logp <- -log10(pstats_sva$P.Value.sva)
range(pstats$logp)
range(pstats$logFC.uncor)

range(pstats_sva$logp)
range(pstats_sva$logFC.sva)

volcanoplots_uncor <- itsar::ggvolcano(pstats, xvar = logFC.uncor,
                   yvar = logp,
                   labelvar = Gene.names.uncor,
                   adjpvalvar = adj.P.Val.uncor,
                   fdr.range.cut = c(0,0.01,0.1,1),
                   high.sig.threshold = 0.001,
                   annotgenes = T, xlimits = c(-2, 2), ylimits = c(0, 16))


volcanoplots_sva <- itsar::ggvolcano(pstats_sva, xvar = logFC.sva,
                   yvar = logp,
                   labelvar = Gene.names.sva,
                   adjpvalvar = adj.P.Val.sva,
                   fdr.range.cut = c(0,0.01,0.1,1),
                   high.sig.threshold = 0.001,
                   annotgenes = T, xlimits = c(-2, 2), ylimits = c(0, 16))



```
 
Volcano plot for uncorrected analysis

```{r fig.width = 5, fig.asp = 1.1}

volcanoplots_uncor$ggplot_volcano

```

Volcano plot for SVA batch corrected analyses:
```{r fig.width = 5, fig.asp = 1.1}

volcanoplots_sva$ggplot_volcano

```

Interactive plotly volcano plots 
```{r fig.width = 5, fig.asp = 1.2}

volcanoplots_uncor$plotly_volcano

```

Volcano plot for SVA batch corrected analyses:
```{r fig.width = 5, fig.asp = 1.2}

volcanoplots_sva$plotly_volcano

```
